Lifting CDCL to Relational Domains

### 2016-10-14 example.tex
experiment reference
example finalised

\section{Motivating Example}
%
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & Control-Flow Graph & Abstract Conflict Graph \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
void main() {
int x,sepc=0,impl=0;
assume(x>=1 && x<=N);
spec = 2*x; 
int state = 1;
int y=x;
while(state!=0) {
 switch(state) {
  case 0: state = 1; 
  break;
  case 1: state = 2; 
  break;
  case 2: x--; y++;
  if(x == 0) state=0;
  else state=1;break;
 } 
}
impl = y;
assert(spec == impl);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
\centering
\vspace*{0.3cm}
\scalebox{.45}{\import{figures/}{example.pspdftex}}
%\caption{A control flow graph for P, P' and abstract conflict graph (ACFG) \label{fig:filter}}
\end{minipage}
&
\begin{minipage}{4.40cm}
\centering
\vspace*{0.3cm}
\scalebox{.45}{\import{figures/}{acdl_run.pspdftex}}
%\caption{A control flow graph for P, P' and abstract conflict graph (ACFG) \label{fig:filter}}
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, CFG and Abstract Conflict Graph}
\label{example}
\end{figure}
Let us consider a simple program in Figure~\ref{example} that checks 
the equivalence of a state-machine based interger multiplication against  
a specification that uses multiplication operator (*).  The assertion 
simply compares the result of two multiplication for an integer input $x$.  
To avoid overflow error, we assume $x$ has an upper bound $N$.  The 
control flow graph is shown in the middle column of Figure~\ref{example}.

We analyze the program using three different analyzers -- {\em bounded 
model checking (BMC)}, {\em abstract interpretation} and {\em ACDCL}.    
For $N=2$ and unwind limit of $bound=4$ for the {\em while} loop,
table~\ref{solver} shows the statistics from MiniSAT~\cite{minisat} solver 
for analysis using BMC.  The SAT solver makes 36 decisions, 1365 boolean 
constraint propagations and performs 5 conflict analysis with 65 conflict 
literals.  Moreover, the solver performs a restart, presumably to recover 
from bad decisions.

On the other hand, abstract interpretation using forward analysis 
is too imprecise to verify the safety of the above program using 
interval domain.  The imprecision is due to control-flow join at 
node $n1$.  State-of-the-art abstract interpretation tool, Astr{\'e}e, 
requires external hints, provided by manually annotating the code 
with partition directives at $n1$, to prove safety.  The partition 
directive helps the tool to analyze each program execution paths 
separately.  To do so, Astr{\'e}e requires a product of trace 
partitioning domain and an interval domain to determine safety.  
Consequently, the analysis becomes very expensive for a large value 
of $N$.  In general, the imprecision is either intended by the tool 
because such high precision analysis is normally not required for 
runtime error analysis or the imprecision is unavoidable due to the 
complexity of the application under analysis.  
%
\begin{table}
\begin{center}
{
\begin{tabular}{l|l|l|l|l|l}
\hline
Solver & Decisions & propagations & conflict & conflict literals & restarts \\ \hline
SAT & 36 & 1365 & 5 & 65 & 1 \\ \hline
ACDCL & 1 & 23 & 1 & 1 & 0 \\
\hline
\end{tabular}
}
\end{center}
\caption{Solver-statistics}
\label{solver}
\end{table}
%
The right-side of Figure~\ref{example} shows an analysis using ACDCL with 
interval domain.  The analysis associates an interval for each control 
location and variable, as shown in the abstract conflict graph.  Interval 
constraints generated from forward analysis in the initial deduction phase 
infers that $\langle x:[-\infty,2], spec:[2,4], impl:[-1,\infty],
state:[1,\infty], y:[1,+\infty] \rangle$.  Clearly, ACDCL can not prove safety.

Hence, ACDCL makes an arbitrary decision $x:[-\infty,1]$ to refine the analysis.  
The decision constrains the interval of $x$.  The deductions made during fixed point
iteration are represented by abstract conflict graph in right-side of 
figure~\ref{example}.   Nodes in the graph denotes the corresponding 
updates to the program variables in the cfg.  Interval analysis concludes 
that the $(n12:\bot)$, that is the decision $x:[-\infty,1]$ leads to a 
{\em conflict}. Intuitively, it means that the program is {\em safe} for 
$x:[-\infty,1]$.

A clause learning SAT solver would learn the reason for conflict at this 
point and then backtrack to a level such that the learnt clause is UNIT.  
Similar to conflict analysis phase in SAT solvers, ACDCL learns that 
(\(I: x = [2,+\infty]\)), that is all error traces must satisfy $(x \geq 2)$ 
at node $I$.  The analysis discards all interval constraints that
lead to the conflict and backtrack to decision level 0.  ACDCL then 
performs interval analysis with the learnt constraint $(x>=2)$ which again 
leads to a conflict.  The analysis cannot backtrack further and therefore 
it terminates proving that the program is safe.  A DPLL-style solver would 
perform proof by cases.  However, decision and clause learning are used 
to avoid case based reasoning and prevents enumeration behavior.

The result of the analysis using ACDCL is shown in table~\ref{solver}.  
Compared to a SAT solver, ACDCL makes only a single decision which lead 
to 23 interval constraint propagations, and learns a single conflict clause 
involving only one literal.  Compared to abstract interpretation, ACDCL does 
not require external hints to proof the program, thus automatically performing 
program and property driven trace partitioning to generate proofs using decision 
and clause learning.

>>>>

\section{Motivating Example}
%
\begin{figure}[t]
\scriptsize
\begin{tabular}{l|l|l}
\hline
C program & Control-Flow Graph & Abstract Conflict Graph \\
\hline
\begin{lstlisting}[mathescape=true,language=C]
void main() {
 int x,z;
 int val;
 assume(val>=0 && val<=N);
 _Bool c;
 if(c)
  x = val;
 else 
  x = -val;
 z = x * x;
 assert(z>=0);
}
\end{lstlisting}
&
\begin{minipage}{4.40cm}
\centering
\vspace*{0.3cm}
\scalebox{.45}{\import{figures/}{example.pspdftex}}
%\caption{A control flow graph for P, P' and abstract conflict graph (ACFG) \label{fig:filter}}
\end{minipage}
&
\begin{minipage}{4.40cm}
\centering
\vspace*{0.3cm}
\scalebox{.45}{\import{figures/}{acdl_run.pspdftex}}
%\caption{A control flow graph for P, P' and abstract conflict graph (ACFG) \label{fig:filter}}
\end{minipage}
\\
\hline
\end{tabular}
\caption{C Program, Control-flow Graph (CFG) and Abstract Conflict Graph}
\label{example}
\end{figure}
Let us consider a simple program in Figure~\ref{example} that performs 
multiplication of two machine integers and checks that the result of the 
multiplication is always positive for an integer input $x$.  To avoid 
overflow error, we assume $x$ has an upper bound $N$.  The control 
flow graph is shown in the middle column of Figure~\ref{example}.

We analyze the program using three different analyzers -- {\em bounded 
model checking (BMC)}, {\em abstract interpretation} and {\em ACDCL}.    
\Omit {
The SAT solver makes 36 decisions, 1365 boolean constraint propagations 
and performs 5 conflict analysis with 65 conflict literals.  Moreover, 
the solver performs a restart, presumably to recover from bad decisions.  
}
Standard BMC tools convert the program to a bit-vector equation and pass 
them to a SAT solver.  However, standard forward interval analysis is 
too imprecise to verify the safety of the above program.  The 
imprecision is due to control-flow join at node $n4$.  State-of-the-art 
abstract interpretation tool, Astr{\'e}e, requires external hints, provided 
by manually annotating the code with partition directives at $n1$, 
to prove safety.  The partition directive helps the tool to analyze 
each program execution paths separately.  To do so, Astr{\'e}e 
requires a product of trace partitioning domain and an interval domain 
to determine safety.  However, a relational domain such as octagons is 
able to prove the program without any partition.  In general, the 
imprecision is either intended by the tool because such high 
precision analysis is normally not required for runtime error 
analysis or the imprecision is unavoidable due to the complexity 
of the application under analysis.  
%
\begin{table}
\begin{center}
{
\begin{tabular}{l|l|l|l|l|l}
\hline
Solver & Decisions & propagations & conflict & conflict literals & restarts \\ \hline
SAT & 233 & 36436 & 162 & 2604 & 2 \\ \hline
ACDCL (Interval) & 1 & 17 & 1 & 1 & 0 \\ \hline
ACDCL (Octagons) & 0 & 7 & 0 & 0 & 0 \\ 
\hline
\end{tabular}
}
\end{center}
\caption{Solver statistics}
\label{solver}
\end{table}
%
The right-side of Figure~\ref{example} shows an analysis using ACDCL with 
octagon domain (top) and an interval domain (bottom).  The analysis 
associates an abstract element for each control location and variable, 
as shown in the abstract conflict graph in Firgure~\ref{example}.  
Octagonal constraints deduced by ACDCL solver are sufficient to prove 
safety without any decision or clause learning.  However, unlike Ast{\'e}e, 
an automatic interval analysis is also sufficient to prove safety with 
only a single decision.  Interval constraints generated from forward 
analysis in the initial deduction phase infers that 
$\langle x:[-5,5], z:[-25,25] \rangle$.  Clearly, ACDCL can not prove safety.

Hence, ACDCL makes a decision $c:=[1,1]$ to refine the analysis.  
ACDCL implements several decision heuritics.  In this case, the 
branching variable $c$ is chosen by an {\em ordered} decision 
heuristics that select variables appearing in conditional branches 
first before choosing other program variables. 
The decision constrains the interval of $x:=[0,5]$.  The deductions made 
during fixed point iteration are represented by abstract conflict 
graph in right-side of figure~\ref{example}.  Nodes in the abstract 
conflict graph denotes the corresponding updates to the program 
variables in the CFG.  Interval analysis concludes that the $(ERROR:\bot)$, that 
is the decision $c:=[1,1]$ leads to a {\em conflict}.  Thus, 
the program is {\em safe} for $c:=[1,1]$.

A clause learning SAT solver would learn the reason for conflict at this 
point and then backtrack to a level such that the learnt clause is UNIT.  
Similar to conflict analysis phase in SAT solvers, ACDCL learns that 
(\(I: c:=[0,0]\)), that is all error traces must satisfy $(c \neq 1)$ 
at node $I$.  The analysis discards all interval constraints that
lead to the conflict and backtrack to decision level 0.  ACDCL then 
performs interval analysis with the learnt constraint $(c \neq 1)$ which also 
leads to a conflict.  The analysis cannot backtrack further and therefore 
it terminates proving that the program is safe.  A DPLL-style solver would 
perform proof by cases.  However, decision and clause learning are used 
to avoid case based reasoning and prevents enumeration behavior.

For $N=46000$, table~\ref{solver} shows the statistics from 
MiniSAT~\cite{minisat} solver for analysis using BMC, and 
analysis using ACDCL with interval and octagon domains.  
Compared to a SAT solver, there is a significant reduction 
in the number of decisions, propagations, learnt clauses and 
restrarts.  Compared to abstract interpretation, ACDCL does 
not require external hints to proof the program, thus 
automatically performing program and property driven trace 
partitioning to generate proofs using decision and clause learning.
